### 2022-02-16 introduction/introduction.tex
more

The biggest contribution of this thesis is the extension of two theorem provers,
E and Zipperposition \cite{sc-15-simon-phd}, to full higher-order logic. To
extend those provers, we faced many challenges. Some of them were of engineering
nature, some of them were algorithmic, while others concerned fine-tuning the
heuristics to achieve best possible performance. We discuss those challenges in more detail:
\begin{itemize}
  \item We implemented the complete superposition calculus for $\lambda$-free
  higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
  We had to modify internal term structure of Ehoh to support $\lambda$-free
  higher-order terms, implement new unification and matching algorithms, extend
  indexing data structures to work with higher-order terms, and to improve the
  performance of heuristics on higher-order problems.
  \item Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second stage described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in short amount of time we decided to
  implement the calculus in Zipperposition. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition with Zipperposition
  \cite{gs-19-casc27}. It took third place, closely behind Leo-III (1 percent
  point) \cite{sb-21-leo3} and far behind (12 percent points) Satallax
  \cite{cb-12-satallax}.
  \item Through this experience we identified one of the bottlenecks of
  Zipperposition: its unification algorithm. To remove this bottleneck we
  developed a complete higher-order unification algorithm that was more
  efficient and less redundant than current state of the art. 
  \item We also noticed that Zipperposition's support for first-class Booleans
  was lacking compared to competition. We have closely studied the problems on
  which Zipperposition fails, but are easily solved by other provers and created
  a set of incomplete rules that enhance Zipperposition's Boolean support.
  Zipperposition 2 implemented these rules together with the new unification
  algorithm. In 2020, Zipperposition 2 was the winner of higher-order division
  of CASC \cite{gs-21-cascj10}. This time it was 20 percent points better than the second best
  prover, Satallax.
  \item Inspired by this success we implemented the complete superposition
  calculus for full higher-order logic (i.e., the third stage) that we developed with Bentkamp and other
  colleagues. The development of complete calculus helped us understand in which
  ways our pragmatic extension was too weak to prove hard theorems of
  higher-order logic. We further tuned the heuristics and implemented new ways
  to tame the search space explosion. In 2021 we released Zipperposition 2.1
  which implemented the new calculus and the described tweaks. It again won
  CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
  runner-up, Vampire 4.5 \cite{lkav-13-vampire}.

  \item Taking all the experience we gathered while working on Zipperposition,
  we further extended Ehoh to full higher-order logic. We again modified term
  structure, implemented new unification and matching algorithms, extended indexing data
  structures, and tweaked heuristics. Even though this extension has larger scope than 
  the original extension from E to Ehoh, strong basis that we have earlier built
  made it manageable.

  \item Following our approach of lifting efficient provers from
  first-order to higher-order logic, we went a step back in an effort to improve
  first-order provers. In particular, we looked at the most efficient approaches
  to simplify problems in propositional logic, a logic that disallows quantification,
  but it is decidable. We lifted some of these approaches to first-order logic
  and implemented them in Zipperposition. 
\end{itemize}

@@

\begin{itemize}
  \item Martin Desharnais, Petar Vukmirović, Jasmin Blanchette, and Makarius Wenzel. Seventeen Provers Under the Hammer.
  Draft paper, submitted to ITP 2022.
  \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, and Petar Vukmirović. Superposition for Higher-Order Logic.
  Draft article, submitted to Journal of automated reasoning.
  \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, Petar Vukmirović, and Uwe Waldmann.  Superposition with Lambdas.
  \emph{Journal of Automated Reasoning} 65(7): 893--940, 2021. 
  \item Visa Nummelin, Alexander Bentkamp, Sophie Tourret, and Petar Vukmirović. Superposition with First-Class Booleans and Inprocessing Clausification.
  In Platzer, A., Sutcliffe, G. (eds.) \emph{CADE-28}, LNCS 12699, pp. 378--395, Springer, 2021.
  \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, and Petar Vukmirović. Superposition for Full Higher-Order Logic.
  In Platzer, A., Sutcliffe, G. (eds.) \emph{CADE-28}, LNCS 12699, pp. 396--412, Springer, 2021.
  \item Stephan Schulz, Simon Cruanes, and Petar Vukmirović. Faster, Higher, Stronger: E 2.3.
  In Fontaine, P. (ed.) \emph{CADE-27},  LNCS 11716, pp. 495--507, Springer, 2019.
  \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, Petar Vukmirović, and Uwe Waldmann.  Superposition with Lambdas.
  In Fontaine, P. (ed.) \emph{CADE-27}, LNCS 11716, pp. 55--73, Springer, 2019. 
\end{itemize}

>>>>

The biggest contribution of this thesis is the extension of two theorem provers,
E and Zipperposition \cite{sc-15-simon-phd}, to full higher-order logic. To
extend those provers, we faced many challenges. Some of them were of engineering
nature, some of them were algorithmic, while others concerned fine-tuning the
heuristics to achieve best possible performance. We discuss those challenges in more detail:
\begin{itemize}
  \item We implemented the complete superposition calculus for $\lambda$-free
  higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
  We had to modify internal term structure of Ehoh to support $\lambda$-free
  higher-order terms, implement new unification and matching algorithms, extend
  indexing data structures to work with higher-order terms, and to improve the
  performance of heuristics on higher-order problems.
  \item Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second stage described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in short amount of time we decided to
  implement the calculus in Zipperposition. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition with Zipperposition
  \cite{gs-19-casc27}. It took third place, closely behind (1 percent
  point) Leo-III \cite{sb-21-leo3}  and far behind (12 percent points) Satallax \cite{cb-12-satallax}.
  \item Through this experience we identified one of the bottlenecks of
  Zipperposition: its unification algorithm. To remove this bottleneck we
  developed a complete higher-order unification algorithm that was more
  efficient and less redundant than current state of the art. 
  \item We also noticed that Zipperposition's support for first-class Booleans
  was lacking compared to competition. We have closely studied the problems on
  which Zipperposition fails, but are easily solved by other provers and created
  a set of incomplete rules that enhance Zipperposition's Boolean support.
  Zipperposition 2 implemented these rules together with the new unification
  algorithm. In 2020, Zipperposition 2 was the winner of higher-order division
  of CASC \cite{gs-21-cascj10}. This time it was 20 percent points better than the second best
  prover, Satallax.
  \item Inspired by this success we implemented the complete superposition
  calculus for higher-order logic with Booleans (i.e., the third stage) that we developed with Bentkamp and other
  colleagues. Developing this calculus helped us understand in which
  ways our pragmatic extension was too weak to prove hard theorems of
  higher-order logic. We further tuned the heuristics and implemented new ways
  to tame the search space explosion. In 2021 we released Zipperposition 2.1
  which implemented the new calculus and the described tweaks. It again won
  CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
  runner-up, Vampire 4.5 \cite{lkav-13-vampire}.

  \item Taking all the experience we gathered while working on Zipperposition,
  we further extended Ehoh to full higher-order logic, obtaining prover we call
  \newehoh{}. We again modified term structure, implemented new unification and
  matching algorithms, extended indexing data structures, and tweaked
  heuristics. Even though this extension has larger scope than the original
  extension from E to Ehoh, strong basis that we have earlier built made it
  manageable.

  \item Following our approach of lifting efficient provers from
  first-order to higher-order logic, we went a step back in an effort to improve
  first-order provers. In particular, we looked at the most efficient approaches
  to simplify problems in propositional logic, a simpler logic that disallows quantification,
  but it is decidable. We lifted some of these approaches to first-order logic
  and implemented them in Zipperposition. 
\end{itemize}

@@

\begin{itemize}
  \item Martin Desharnais, Petar Vukmirović, Jasmin Blanchette, and Makarius Wenzel. Seventeen Provers Under the Hammer.
  Draft paper.
  \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, and Petar Vukmirović. Superposition for Higher-Order Logic.
  Draft article.
  \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, Petar Vukmirović, and Uwe Waldmann.  Superposition with Lambdas.
  \emph{Journal of Automated Reasoning} 65(7): 893--940, 2021. 
  \item Visa Nummelin, Alexander Bentkamp, Sophie Tourret, and Petar Vukmirović. Superposition with First-Class Booleans and Inprocessing Clausification.
  In Platzer, A., Sutcliffe, G. (eds.) \emph{CADE-28}, LNCS 12699, pp. 378--395, Springer, 2021.
  \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, and Petar Vukmirović. Superposition for Full Higher-Order Logic.
  In Platzer, A., Sutcliffe, G. (eds.) \emph{CADE-28}, LNCS 12699, pp. 396--412, Springer, 2021.
  \item Stephan Schulz, Simon Cruanes, and Petar Vukmirović. Faster, Higher, Stronger: E 2.3.
  In Fontaine, P. (ed.) \emph{CADE-27},  LNCS 11716, pp. 495--507, Springer, 2019.
  \item Alexander Bentkamp, Jasmin Blanchette, Sophie Tourret, Petar Vukmirović, and Uwe Waldmann.  Superposition with Lambdas.
  In Fontaine, P. (ed.) \emph{CADE-27}, LNCS 11716, pp. 55--73, Springer, 2019. 
\end{itemize}
